Predictive Modeling of I/O Characteristics \\ in High Performance Computing Systems

### 2017-11-12 HPC_Paper/paper.tex
Working on paper, updated CSV result files to have decimal points in all predicted columns, checked function of data collection for SVR polynomial.

Each of high performance computing, cloud computing, and computer
security have their own interests in modeling and predicting the
performance of computers with respect to how they are configured. An
effective model might infer internal mechanics, minimize power
consumption, or maximize computational throughput of a given
system. This paper analyzes a seven-dimensional dataset measuring the
input/output (I/O) characteristics of a cluster of identical computers
using the benchmark IOzone. The I/O performance characteristics are
modeled with respect to system configuration using multivariate
interpolation and approximation techniques. The analysis reveals that
accurate models of I/O characteristics for a computer system may be
created from a small fraction of possible configurations, and that
some modeling techniques will continue to perform well as the number
of system parameters being modeled increases. These results have
strong implications for future predictive analyses based on more
comprehensive sets of system parameters.

@@

Many previous works attempting to model system performance have used
simulated environments to estimate the performance of a
system. \cite{grobelny2007fase} \cite{wang2009simulation}
\cite{wang2013towards} Some of these works cite statistical models as
being over-simplified and not capabel of capturing the true complexity
of the underlying system. This claim is partially correct, noticing
that a large portion of predictive statistical models rely on
simplifying the system to one or two
parameters. \cite{snavely2002framework} \cite{bailey2005performance}
\cite{barker2009using} \cite{ye2010analyzing} These limited
statistical models have provided satisfactory performance in very
narrow application settings. Many of the aforementioned statistical
modeling techniques claim to generalize, while simultaneously
requiring additional code annotations, hardware abstractions, or
additional application level understandings in order to generate
models. The approach presented here requires no modifications of the
application, no architecture abstractions, nor any structural
descriptions of the input data being modelled. The techniques used are
purely mathematical and only need formatted data as input.

Among the statistcal models presented in prior works,
\cite{bailey2005performance} specifically mentioned that it is
difficult for the models to capture variability introduced by
I/O. System variability in general has become a problem of increasing
interest to the systems and HPC communities, however most of the work
has focused on Operating System induced variability.
\cite{beckman2008benchmarking} \cite{de2007identifying} The work that
has focused on managing I/O variability, \cite{lofstead2010managing},
does not use any sophisticated modeling techniques. Hence, this paper
presents a case study applying advanced mathematical and statistical
modeling techniques to the domain of HPC system I/O
characteristics. The models are used to predict the mean throughput of
a system and the variance in throughput of a system. The discussion
section outlines how the exact techniques presented can be applied to
any performance metric and any system.

@@

In general, this paper compares five multivariate approximation
techniques that operate on inputs in $\mathbb{R}^d$ (vectors of $d$
real numbers) and produce predicted responses in $\mathbb{R}^1$. In
order to provide coverage of the varied mathematical strategies that
can be employed to solve the continuous modeling problem, three of the
techniques are regression-based and the remaining two techniques
produce interpolants. The sections below outline the mathematical
formulations of each technique and provide computational complexity
bounds with respect to the size (number of points and dimension) of
input data.

\subsection{Approximation}
Multivariate approximations are capable of accurately modelling
complex relationships between the dimensions of any set of points in
$\mathbb{R}^n$ with respect to some response in $\mathbb{R}$. The
approximations produce a function $f: \mathbb{R}^n \rightarrow
\mathbb{R}$ which minimizes some metric related to the provided data.

\subsubsection{Multivariate Adaptive Regression Splines}
\subsubsection{Multi-Layer Perceptron Regressor}
\subsubsection{Support Vector Regressor}

>>>>

Each of high performance computing, cloud computing, and computer
security have their own interests in modeling and predicting the
performance of computers with respect to how they are configured. An
effective model might infer internal mechanics, minimize power
consumption, or maximize computational throughput of a given
system. This paper analyzes a four-dimensional dataset measuring the
input/output (I/O) characteristics of a cluster of identical computers
using the benchmark IOzone. The I/O performance characteristics are
modeled with respect to system configuration using multivariate
interpolation and approximation techniques. The analysis reveals that
accurate models of I/O characteristics for a computer system may be
created from a small fraction of possible configurations, and that
some modeling techniques will continue to perform well as the number
of system parameters being modeled increases. These results have
strong implications for future predictive analyses based on more
comprehensive sets of system parameters.

@@

Many previous works attempting to model system performance have used
simulated environments to estimate the performance of a
system. \cite{grobelny2007fase,wang2009simulation,wang2013towards}
Some of these works cite statistical models as being over-simplified
and not capabel of capturing the true complexity of the underlying
system. This claim is partially correct, noting that a large portion
of predictive statistical models rely on simplifying the system to one
or two parameters.
\cite{snavely2002framework,bailey2005performance,barker2009using,ye2010analyzing}
These limited statistical models have provided satisfactory
performance in very narrow application settings. Many of the
aforementioned statistical modeling techniques claim to generalize,
while simultaneously requiring additional code annotations, hardware
abstractions, or additional application level understandings in order
to generate models. The approach presented here requires no
modifications of the application, no architecture abstractions, nor
any structural descriptions of the input data being modelled. The
techniques used are purely mathematical and only need formatted data
as input.

Among the statistcal models presented in prior works,
\cite{bailey2005performance} specifically mentions that it is
difficult for the simplified models to capture variability introduced
by I/O. System variability in general has become a problem of
increasing interest to the HPC and systems communities, however most
of the work has focused on Operating System (OS) induced variability.
\cite{beckman2008benchmarking,de2007identifying} The work that
has focused on managing I/O variability, \cite{lofstead2010managing},
does not use any sophisticated modeling techniques. Hence, this paper
presents a case study applying advanced mathematical and statistical
modeling techniques to the domain of HPC I/O characteristics. The
models are used to predict the mean throughput of a system and the
variance in throughput of a system. The discussion section outlines
how the exact techniques presented can be applied to any performance
metric and any system.

@@

In general, this paper compares five multivariate approximation
techniques that operate on inputs in $\mathbb{R}^d$ (vectors of $d$
real numbers) and produce predicted responses in $\mathbb{R}$. In
order to provide coverage of the varied mathematical strategies that
can be employed to solve the continuous modeling problem, three of the
techniques are regression-based and the remaining two techniques
produce interpolants. The sections below outline the mathematical
formulations of each technique and provide computational complexity
bounds with respect to the size (number of points and dimension) of
input data. Throughout the sections, $d$ will refer to the dimension
of the input data, $n$ is the number of points in the input data,
$x_i \in \mathbb{R}^d$ is the $i^{th}$ input data point, and $y_i$ is
the response value of the $i^{th}$ input data point.

\subsection{Approximation}
Multivariate approximations are capable of accurately modelling
complex relationships between the dimensions of any set of points in
$\mathbb{R}^d$ with respect to some response in $\mathbb{R}$. The
approximations produce a function $f: \mathbb{R}^d \rightarrow
\mathbb{R}$ which minimizes some selected error metric related to the
provided data. Often, as in least-squares regression, the distance
metric chosen is the sum of squared differences between modeled
response values and true response values.

\subsubsection{Multivariate Adaptive Regression Splines}
This algorithm was introduced by statistician J. Friedman in
\cite{friedman1991multivariate} and subsequently improved to it's
current version in \cite{stanford1993fast}. This work uses the Fast
Multivariate Adaptive Regression Splines (Fast MARS) formulation. In
Fast MARS, a least-squares fit model is iteratively built by beginning
with a single constant valued function and adding two new basis
functions at each iteration of the form

$$ B_{2s-1}(x) = B_l(x) [c(x_i-v)]_+ $$
$$ B_{2s}(x) = B_k(x) [c(x_i-v)]_- $$

Where $s$ is the iteration number, $B_l(x)$ and $B_k(x)$ are some
previously generated basis functions, $c, v \in \mathbb{R}$, and $x_i$
with $0 < i \leq d$ is the $i^{th}$ component (1 indexed) of input
vector $x$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise linear
approximation of the data that is $C^0$. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is a user
selected value that is the maximum complexity of the model in terms of
the number of underlying basis functions. This paper uses a default
max number of basis functions of $n$, however with larger datasets ($n
> 1000$)this value would need to be reduced.

\subsubsection{Multi-Layer Perceptron Regressor}
The neural network is a well studied and widely used mathematical tool
for both regression and classification tasks.
\cite{hornik1989multilayer} When using the rectified linear unit
(ReLU) activation functions \cite{dahl2013improving} and trained using
the BFGS technique \cite{moller1993scaled}, the model built by a
multi-layer perceptron uses layers of the form

$$l(x) : \mathbb{R}^{i} \rightarrow \mathbb{R}^{j} = \text{max}\big( x
W_l, 0 \big)$$

Where $W_l$ is the $i$ by $j$ weight matrix of layer $l$ and the max
function is applied element-wise to the resulting vector. In this
form, the multi-layer perceptron (MLP) produces a locally linear model
of the input data that is $C^0$. The computational complexity of
training a multi-layer perceptron is approximately $\mathcal{O}(n d
m)$ where $m$ is a constant determined by a linear function of the
sizes of the layers of the network and the convergence stopping
criteria of the BFGS minimization used for finding weights. This paper
uses a single hidden layer with 100 nodes, ReLU activation, and BFGS
for training.

\subsubsection{Support Vector Regressor}
Support vector machines are a common tool used in machine learning
classification tasks that can be adapted for the purpose of regression
\cite{basak2007support}. The mathematical formalisms for exactly how
to build a support vector regressor (SVR) are beyond the scope of this
summary, but the core of the technique solves an optimization problem
of the form

$$ \text{Minimize } \bigg\{ \frac{1}{2}\sum_{i=1}^{n}\sum_{j=1}^{n}
a_i a_j K(x_i, x_j) + \epsilon \sum_{i=1}^{n} a_i - \sum_{i=1}^{n} y_i
a_i \bigg\} $$

$$ \text{Subject to } \sum_{i=1}^{n}a_i = 0 \text{ } \text{ and }
\text{ } a_i \in [0,C] $$

Where $K$ is the selected kernel function, $a \in \mathbb{R}^n$ is the
solved coefficient vector, and $\epsilon$ is the error tolerance.
